---
title: "SummarizingData_filled"
author: "Amy King-Henry"
date modified: "September 2, 2026"
format: html
editor: visual
---

# Miscellaneous Data Summarizing Skills

### Components of a boxplot

- The middle line is the **median**.

- The box is the **interquartile range, or IQR** - the bounds of the second and third quartiles.

- The upper whisker extends from the hinge to the **largest value no further than 1.5 \* IQR** from the hinge (where IQR is the inter-quartile range, or distance between the first and third quartiles).

- The lower whisker extends from the hinge to the **smallest value at most 1.5 \* IQR** of the hinge.

- Data beyond the end of the whiskers are called "outlying" points and are plotted individually.

These are NOT mean and standard deviation!

### Summary Statistics for Continuous Variables

```{r, summary_stat_functions}
# This chunk will not run, list of functions!!!!! 
mean() : Mean
min() : minimum value
max() : maximum value
range() : min and max values
var() : Variance
sd() : Standard deviation
sd()/mean() : Coefficient of Variation
sd()/sqrt(n()) : Standard error
n() : sample size/count of rows
median() : Median
IQR() : Interquartile Range
quantile() : quantiles
summary() : min, 1st quartile, median, mean, 3rd quartile, max
```

```{r}
load(file = "Demo Data Files/CCdata_tidy.Rdata")
install.packages("car")
library(tidyverse)
```

### Example using summarize to make a nice summary table of some old owl limpet data:

```{r}
Size_summary <- CCdata_tidy |>  
  group_by(Impact, Year, Season, Survey_Date) |>  #grouping for everything except                       size  
  summarize(
           mean_size = mean(Size, na.rm = TRUE),  # calculates mean, na.rm remvoes                    the NAs and makes sure the calculation goes through 
            sd_size = sd(Size, na.rm = TRUE),     # standard deviation
            n = n(),                              # count
            se_size = sd_size/sqrt(n)) |>         # standard error
  
  
  mutate_at(vars(mean_size, sd_size, se_size),     # bonus line to reduce digits
            .funs = round, 2)

Size_summary |>                                   #now we're putting the summary                         stats, which we defined before into a new set of functions 
  select( -Survey_Date) |>                        #selecting everything EXCEPT                                                      survey date 
  knitr::kable()                                  #makes nice looking table with                                                    knitr

```

#### Example creating a plot from that summary data:

```{r}
#turning summary table into plot: 
Size_summary |> 
  ggplot() +       #piping into ggplot, leave it blank 
  aes(x = as.factor(Survey_Date) , y = mean_size , fill = Impact) +  #setting up                          axis and color distinctions  
# geom_point(position = position_dodge(0.9)) 
  geom_col(position = position_dodge())  +         #column plot with position dodge
  
  geom_errorbar(aes (ymin = mean_size - se_size, ymax = mean_size + se_size),         position = position_dodge(0.9), width = 0.5) #sets up error bar parameters, needs a new set of aesthetics 

#ymin is the mean size minus the standard erorr size, error bar that goes down --> YMIN!!! 
#ymax is topmost error bar, this is the mean plus the standard error size 
#width in error bar is self explanatory 
#set position dodge to 0.9 when in doubt 
```

```{r}
Size_summary |> 
  ggplot() +       #piping into ggplot, leave it blank 
  aes(x = as.factor(Survey_Date) , y = mean_size , color = Impact) +  #setting up                          axis and color distinctions, distinguishing by impact level                         of site   
  geom_point(position = position_dodge(0.9)) + 

  geom_errorbar(aes (ymin = mean_size - se_size, ymax = mean_size + se_size),         position = position_dodge(0.9), width = 0.5)
```

#### Beyond the summary stats: frequency distributions

Boiling it down to the mean and standard deviation really loses a lot of information though. It's always good to look at the histogram/distribution to get a better sense of what's going on. Let's look back at just that one histogram.

```{r}
CC_sp19 <- CCdata_tidy |> 
  filter(Season == "Spring" & Year == 2019 & Impact == "High" ) #filtering for only the high impact, in 2019, in spring. ands are to meet all 3 conditions 

CC_sp19 |> 
  ggplot() + 
  aes (x = Size) + 
  geom_density()
  # geom_histogram(aes(y = after_stat(density))) #instead of having a count on the y axis, its now a probability density!! the area underneath all of these bars is now equal to 1 

#coord_flip  --> flips the axes 


#there is a minimum cutoff for limpet size -- data artifact! this makes it so that the distribution is skewed towards smaller sizes, even though it could be expected to be symmetrical 
```

What observations can we make about the distribution of this data?

- Is it symmetrical or skewed?

- Are there many outliers?

- Is there missing data?

- Is it unimodal, i.e., is there only one peak where there are the most observations, or are there multiple peaks (bimodal, etc.)?

- What are our expectations for what size demographics should look like for populations of:

  - long-lived vs. short-lived animals

    - owl limpets are long-lived, so this looks right with more older individuals

  - r vs. K-selected

  - growing vs. shrinking populations?

- What factors apart from the population itself may have influenced this pattern? Let's look at the histograms for all the dates and sites:

```{r many histograms}
CCdata_tidy |> 
  ggplot() +
  aes(x = Size) + 
  geom_histogram() + 
  facet_wrap(Survey_Date ~ Impact) 

#remember, facet wraps takes each possible combo of the selected variables 




  


```

On the whole, how would you describe these distributions?

#### The Normal Distribution

Many statistical tests and linear models are based on assumptions that data is normally distributed! Therefore, one thing we'll do a lot of is assess distributions and decide if they can be reasonably approximated by the normal distribution or not.

This applies only to data within a sample, but in the context of a statistical test we will apply tests of normality to the residuals, or the differences from the observed and expected values.

How do we assess?

1.  Visually. Is it symmetrical? Are most of the observations in the middle?

2.  With a QQ plot. Do the observations mostly fall close to the 1:1 line?

3.  With a Shapiro-Wilk test (we'll learn this later)

Visual:

```{r}


CC_sp19 |> 
  ggplot() +
  aes(x = Size) + 
  geom_histogram(aes(y = after_stat(density)))
  
```

2.  With a QQ plot

```{r}
# here's one function from Base R
qqnorm(CC_sp19 $ Size)  #put data into function, $ selects only one column 

# And another from the car package that has 95% confidence intervals
car::qqPlot(CC_sp19 $ Size)  #includes 95% confidence intervals with the car package 
```

Next week, we will start learning about statistical tests which can quantify and assign a probability to how closely your data fits a particular type of distribution.

If they are normal, then we can use a class of tests called parametric tests. If not, we use non-parametric tests.

## Assessing a MODEL for normality

It's one thing to assess a single sample for normality, but when we do a statistical test, we don't actually worry too much about the data in the samples themselves - what matters is the normality of the residuals, or the individual differences of Observed - Expected.

Here is an example dataset of masses of organisms taken from two populations.

```{r}
mass_g1 <- exp()
mass_g2 <- e
mass_g <- c()

fakedata <- tibble(Group = )

```

We ultimately plan to do a t-test, but the easiest way to assess for normality is to make a simple linear model and make some automatic residual plots.

```{r}
model <- lm()
plot()

```

The data itself also does look skewed, but it's important that we view it separately for the two samples:

```{r}
ggplot() + 
  aes(x = , y = ) + 


# the variance is really skewed! 
```

```{r}
ggplot() +  
  facet_wrap()
```

Maybe a log transformation will help!

To log transform this data, we can create a new column or just apply the transformation in the ggplot or model.

```{r}
fakedata <-  |> 
  mutate()

log_model <- lm()

plot()

```

Those diagnostic plots look much better!

The data look more like normal histograms when you plot the samples separately as well.

```{r}
ggplot() + 
  aes() + 
  geom_histogram() + 
  facet_wrap()
```
